---
title: "Assignment 2"
output:
  html_document: default
  pdf_document: default
---

Load the prerequisites 
```{r include=FALSE}
# -------------------------------------------------------------------------
# 0. Robust Package Installation and Loading
# -------------------------------------------------------------------------

# 1. Define the list of packages needed for the analysis
required_packages <- c(
  "tidyverse", # For data manipulation and plotting
  "haven",     # To read SAS data
  "geepack",   # For GEE models
  "lme4",      # For GLMMs
  "emmeans",   # For predicted probabilities (LSMeans)
  "broom",
  "mice" 
)

# 2. Loop through the list: Check -> Install (if needed) -> Load
for (pkg in required_packages) {
  # require() returns TRUE if installed, FALSE if not
  if (!require(pkg, character.only = TRUE)) { 
    message(paste("Installing missing package:", pkg))
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

message("All required packages are installed and loaded.")
```


```{r}
# 0. Load Data
# Assuming the SAS dataset is saved locally
alz_wide <- read_sas("./dataset/alzheimer25.sas7bdat")


# 1. Construct long-format dataset
alz_long2 <- alz_wide %>%
    mutate(
        abpet_0 = abpet0,
        taupet_0 = taupet0
    ) %>%
  # Convert wide to long using regex to capture variable name and time
  pivot_longer(
    cols = matches("(bprs|cdrsb|abpet|taupet)[0-6]"),
    names_to = c(".value", "time"),
    names_pattern = "(.*)(\\d+)" 
  ) %>%
  mutate(
    time = as.numeric(time),
    # Convert key categorical variables to factors immediately
    patid = as.factor(patid),
    sex = as.factor(sex),
    abpet0 = abpet_0,
    taupet0 = taupet_0
    ) %>%
  # Keep specific columns (dplyr::select)
  select(patid, trial, sex, age, edu, bmi, inkomen, job, adl, wzc,
         abpet0, taupet0, abpet, taupet, time, BPRS = bprs, CDRSB = cdrsb)

# View structure
str(alz_long2)
```
# alz_long_all with taupet, abpet etc 
```{r}
# 1. Construct long-format dataset
alz_long_all <- alz_wide %>%
  # Convert wide to long using regex to capture variable name and time
  pivot_longer(
    cols = matches("(bprs|cdrsb|abpet|taupet)[0-6]"),
    names_to = c(".value", "time"),
    names_pattern = "(.*)(\\d+)" 
  ) %>%
  mutate(
    time = as.numeric(time),
    # Convert key categorical variables to factors immediately
    patid = as.factor(patid),
    sex = as.factor(sex), 
    trial = as.factor(trial)
  ) %>%
  # Keep specific columns (dplyr::select)
  select(patid, trial, sex, age, edu, bmi, inkomen, job, adl, wzc,
         ABPET = abpet, TAUPET = taupet, time, BPRS = bprs, CDRSB = cdrsb)



# Create binary outcome for long data
alz_long_all_bin <- alz_long_all %>%
  # Drop missing outcomes
  # Binary outcome: 1 = severe (>10), 0 = non-severe
  mutate(
    CDRSBbin = ifelse(CDRSB > 10, 1, 0),
  )

```



Explore Missingness
```{r}
# 2.1 Create missingness indicator
alz_long2_miss <- alz_long2 %>%
  mutate(missing_cdrsb = ifelse(is.na(CDRSB), 1, 0))

# 2.2 Frequency table (Time * Missing)
table(alz_long2_miss$time, alz_long2_miss$missing_cdrsb)

# 2.3 Proportion missing per time
miss_summary <- alz_long2_miss %>%
  group_by(time) %>%
  summarise(prop_missing = mean(missing_cdrsb))

# Plot: Proportion missing
ggplot(miss_summary, aes(x = time, y = prop_missing)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(title = "Proportion of missing CDRSB by visit year",
       y = "Proportion missing CDRSB", x = "Visit year") +
  theme_minimal()
```


Create binary outcome
```{r}
# Create binary outcome for long data
alz_long2_bin <- alz_long2 %>%
  # Drop missing outcomes
  # Binary outcome: 1 = severe (>10), 0 = non-severe
  mutate(
    CDRSBbin = ifelse(CDRSB > 10, 1, 0),
    time_num = time # time is already numeric from step 1
  )

# Now for wide data 
alz_wide_bin <- alz_wide %>%
  mutate(
    across(starts_with("cdrsb"), ~ifelse(. > 10, 1, 0)),
  )
# Check distribution
table(alz_long2_bin$CDRSBbin, alz_long2_bin$time_num)
```


Descriptive plot
```{r}
# 4.1 Compute proportion severe
prop_severe <- alz_long2_bin %>%
  group_by(time_num) %>%
  summarise(prop_1 = mean(CDRSBbin))

# 4.2 Plot
ggplot(prop_severe, aes(x = time_num, y = prop_1)) +
  geom_col(fill = "darkred") +
  geom_text(aes(label = scales::percent(prop_1, accuracy = 0.1)), vjust = -0.5) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(title = "Proportion of severe CDRSB (CDRSBbin=1 ; >10) by visit year",
       y = "Proportion with CDRSBbin = 1", x = "Visit year") +
  theme_minimal()
```


# Marginal models
## Additive Models + find best working correlation structure
```{r}
# 5.1 Base GEE model: Exchangeable correlation
gee_base <- geeglm(CDRSBbin ~ time_num + age + sex + bmi,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")

summary(gee_base)

# Calculate Odds Ratios (Exponentiate coefficients)
exp(coef(gee_base))


# 5.2 Full GEE model: Exchangeable
gee_full <- geeglm(CDRSBbin ~ time_num + age + sex + bmi + adl + abpet0 + taupet0,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")

summary(gee_full)
exp(coef(gee_full)) # ORs

# 5.2b Full GEE model with timevarying biomarkers: Exchangeable
gee_full_fullvars <- geeglm(CDRSBbin ~ time_num + age + sex + bmi + adl + abpet + taupet,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")

summary(gee_full_fullvars)
exp(coef(gee_full_fullvars)) # ORs


```
## Sensitivity analysis (investigate potential impact of MNAR)
```{r}
# Assuming we want to impute 'cdrsb' at different time points (cdrsb1, cdrsb2, etc.)
cols_to_keep <- c("trial","age", "sex", "bmi","adl", "abpet0", "taupet0", paste0("cdrsb", 0:6))
wide_data <- alz_wide_bin[, cols_to_keep]

# Ensure categorical variables are factors
wide_data$trial <- as.factor(wide_data$trial) 

# --- 2. Shift Adjustment (Sensitivity Analysis) ---
# SAS Equivalent: mnar adjust (diff12 / shift=10 adjustobs=(treat='2'));
# Scenario: What if patients in trial group '1' (or '2') had worse outcomes than predicted?

# a. Perform standard imputation first (MAR assumption)
# We impute cdrsb1-cdrsb6. cdrsb0 is baseline (usually complete).
imp <- mice(wide_data, m = 10, method = "pmm", seed = 12345, printFlag = FALSE)

# b. Apply shift to imputed values
# Let's say we want to add a shift of 2 units to 'cdrsb6' for the treatment group '1'
# This simulates a scenario where dropouts in the treatment group were actually worse off.

imp_shifted <- imp
long_completed_data <- complete(imp, action = "long", include = TRUE)

# Define the shift
shift_value <- 2
target_group <- 1 # Assuming '1' is the treatment group of interest
target_var <- "cdrsb6"

# Apply shift ONLY to imputed values (where .imp > 0) for the specific group
is_missing <- is.na(wide_data[[target_var]])
# We need to map this missingness mask to the long format. 
# Since 'complete' stacks datasets, the row order is preserved per imputation.
n <- nrow(wide_data)
missing_mask <- rep(is_missing, times = (imp$m + 1)) # +1 for original data

# Apply logic: If it was missing AND it's the target group -> add shift
condition <- missing_mask & (long_completed_data$trial == target_group) & (long_completed_data$.imp > 0)

long_completed_data[[target_var]][condition] <- long_completed_data[[target_var]][condition] + shift_value

imp_shifted_mids <- as.mids(long_completed_data)

# c. Analyze the shifted data
# This model analyzes the outcome at the last time point (cdrsb6).
fit_shifted_long <- with(imp_shifted_mids, {
  # 1. Convert to long format within the 'with' call
  long_dat <- alz_wide_bin %>%
    pivot_longer(
      cols = matches("cdrsb[0-6]"),
      names_to = "time_var",
      values_to = "CDRSB_score",
      names_pattern = "cdrsb(\\d+)"
    ) %>%
    mutate(
      time = as.numeric(time_var),
    )
  
  # 2. Run GEE on the long data
  geeglm(cdrsb6 ~ time + age + sex + bmi + adl + abpet0 + taupet0,
         id = patid,
         data = long_dat,
         family = binomial,
         corstr = "exchangeable")
})

pooled_shifted_long <- pool(fit_shifted_long)
print(summary(pooled_shifted_long))

long_dat = complete(imp_shifted_mids, action = "long")
fit_shifted <- with(imp_shifted_mids, geeglm(cdrsb6 ~ age + sex + bmi + adl + abpet0 + taupet0,
                   id = patid,
                   data = imp_shifted_mids["data"],
                   family = binomial,
                   corstr = "exchangeable"))
pooled_shifted <- pool(fit_shifted)
print(summary(pooled_shifted))
```

## Multiplicative Model
```{r}
# 5.4 Full GEE model with Interactions (Exchangeable)
gee_mult <- geeglm(CDRSBbin ~ time_num * ( adl + abpet0 + taupet0) + age + sex + bmi,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")
summary(gee_mult)
exp(coef(gee_mult))

# 5.4 Full GEE model with Interactions (Exchangeable)
gee_mult_fullvars <- geeglm(CDRSBbin ~ time_num * ( adl + abpet + taupet) + age + sex + bmi,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")
summary(gee_mult_fullvars)
exp(coef(gee_mult_fullvars))

# 5.5 Full GEE model with only taupet Interaction (Exchangeable)
gee_mult_taupet <- geeglm(CDRSBbin ~ time_num * taupet + adl + abpet + age + sex + bmi,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")
summary(gee_mult_taupet)
exp(coef(gee_mult_taupet))
```
```{r}
# 5.6 Qaudratic time effects? 
## Full GEE model (Exchangeable)
gee_quadr <- geeglm(CDRSBbin ~ time_num + I(time_num^2) + taupet + adl + abpet + age + sex + bmi,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")
summary(gee_quadr)
exp(coef(gee_quadr))

# 5.6 Qaudratic time effects? 
## Full GEE model (Exchangeable)
gee_mult_quadr <- geeglm(CDRSBbin ~ time_num + I(time_num^2) * taupet + adl + abpet + age + sex + bmi,
                   id = patid,
                   data = alz_long2_bin,
                   family = binomial,
                   corstr = "exchangeable")
summary(gee_mult_quadr)
exp(coef(gee_mult_quadr))
```


lets look at the data without excluding missing outcomes
```{r}
alz_long3_bin <- alz_long2 %>%
  # Binary outcome: 1 = severe (>10), 0 = non-severe
  mutate(
    CDRSBbin = ifelse(CDRSB > 10, 1, 0),
    time_num = time # time is already numeric from step 1
  )

```


# Random effects models
```{r}
# 6.1 Base GLMM
glmm_base <- glmer(CDRSBbin ~ time_num + age + sex + bmi + (1 | patid),
                   data = alz_long2_bin,
                   family = binomial(link = "logit"),
                   nAGQ = 1) # Laplace approximation

summary(glmm_base)


# 6.2 Full GLMM
glmm_full <- glmer(CDRSBbin ~ time_num + age + sex + bmi + adl + abpet0 + taupet0 + (1 | patid),
                   data = alz_long2_bin,
                   family = binomial(link = "logit"),
                   nAGQ = 1)

summary(glmm_full)
```




# Random effects structure check

```{r}


# Create a copy of your data to avoid overwriting the original
alz_long_scaled <- alz_long_all_bin

# --- Step 1: Rescale and Center Continuous Predictors ---
# We apply the scale() function to each continuous predictor.
# Note: We do not scale the outcome 'CDRSBbin', the grouping variable 'patid',
# or categorical predictors like 'sex'.
vars_to_scale <- c("time", "age", "bmi", "adl", "ABPET", "TAUPET", "BPRS")
alz_long_scaled[vars_to_scale] <- lapply(alz_long_scaled[vars_to_scale], scale)


```

```{r}
# Model 1: Random intercept and slope for patid nested within trial
model1 <- glmer(CDRSBbin ~ time + age + sex + bmi + adl + ABPET + TAUPET + BPRS + (time | trial/patid),
                data = alz_long_scaled,
                family = binomial(link = "logit"),
                nAGQ = 1,
                control = glmerControl(optimizer = "bobyqa"))

# Model 2: Random intercept and slope for patid, one random intercept for trial
model2 <- glmer(CDRSBbin ~ time + age + sex + bmi + adl + ABPET + TAUPET + BPRS + (time | patid) + (1 | trial),
                data = alz_long_scaled,
                family = binomial(link = "logit"),
                nAGQ = 1,
                control = glmerControl(optimizer = "bobyqa"))

# Model 3: A random intercept for patid, a random intercept for trial
model3 <- glmer(CDRSBbin ~ time + age + sex + bmi + adl + ABPET + TAUPET + BPRS + (1 | patid) + (1 | trial),
                data = alz_long_scaled,
                family = binomial(link = "logit"),
                nAGQ = 1,
                control = glmerControl(optimizer = "bobyqa"))

# Model 4: A random intercept and slope for patid, no random effects for trial
model4 <- glmer(CDRSBbin ~ time + age + sex + bmi + adl + ABPET + TAUPET + BPRS + (time | patid),
                data = alz_long_scaled,
                family = binomial(link = "logit"),
                nAGQ = 1,
                control = glmerControl(optimizer = "bobyqa"))

# Model 5: A random intercept and slope for patid, no random effects for trial
model5 <- glmer(CDRSBbin ~ time + age + sex + bmi + adl + ABPET + TAUPET + BPRS + (1 | patid),
                data = alz_long_scaled,
                family = binomial(link = "logit"),
                nAGQ = 1,
                control = glmerControl(optimizer = "bobyqa"))

# Create a list of the models
models <- list(model1, model2, model3, model4, model5)

# Set names for the models for easier identification
model_names <- c(
  "Random Intercept & Slope for trial/patid",
  "Random Intercept & Slope for patid + Random Intercept for trial",
  "Random Intercept for patid + Random Intercept for trial",
  "Random Intercept & Slope for patid", 
  "Random Intercept for patid"
)

# Create a data frame to store the results
comparison_table <- data.frame(
  Model = model_names,
  AIC = sapply(models, AIC),
  BIC = sapply(models, BIC)
)

# Print the comparison table
print(comparison_table)
```

```{r}
summary(model4)
```


# Number of ADQ points check 

```{r}
# Define the nAGQ values to test
nAGQ_values <- c(1, 10, 25, 50)

# Create an empty list to store the results
results_list <- list()

# Loop through the nAGQ values
for (n in nAGQ_values) {
  # Fit the glmer model
  model <- glmer(CDRSBbin ~ time + age + sex + bmi + adl + ABPET + TAUPET + BPRS + (1 | patid),
                 data = alz_long_scaled,
                 family = binomial(link = "logit"),
                 nAGQ = n,
                 control = glmerControl(optimizer = "bobyqa"))
  
  # Extract the coefficients and standard errors
  coeffs <- summary(model)$coefficients
  
  # Store the results in the list
  results_list[[as.character(n)]] <- coeffs
}

# Combine the results into a single table
# Get the coefficient names from the first model
all_coeffs <- rownames(results_list[['1']])

# Create a final results table
final_results <- data.frame(row.names = all_coeffs)

# Populate the table with estimates and standard errors from each model
for (n in as.character(nAGQ_values)) {
  final_results[paste0("Est_nAGQ", n)] <- results_list[[n]][, "Estimate"]
  final_results[paste0("SE_nAGQ", n)] <- results_list[[n]][, "Std. Error"]
}

# Print the final results table
print(final_results)
```



Empirical BE Random Intercepts
```{r}
# 7.1 Extract Random Effects
re_df <- ranef(glmm_full)$patid
names(re_df) <- "EB_Intercept"
re_df$patid <- rownames(re_df) # Make ID a column

# 7.2 Distribution Histograms and QQ Plots
# Histogram
ggplot(re_df, aes(x = EB_Intercept)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightgrey", color = "black") +
  stat_function(fun = dnorm, args = list(mean = mean(re_df$EB_Intercept), sd = sd(re_df$EB_Intercept)), color = "red") +
  labs(title = "Empirical Bayes random intercepts (GLMM)") +
  theme_minimal()

# QQ Plot
ggplot(re_df, aes(sample = EB_Intercept)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "QQ Plot of Random Intercepts")


# 7.3 Relationship between EB intercept and Baseline Age
# Merge random effects back with baseline data (time == 0)
baseline_data <- alz_long2_bin %>%
  filter(time_num == 0) %>%
  select(patid, age)

eb_merged <- left_join(re_df, baseline_data, by = "patid")

# Scatter plot with Loess
ggplot(eb_merged, aes(x = age, y = EB_Intercept)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(y = "EB random intercept (log-odds)", x = "Age at baseline",
       title = "EB random intercepts vs Age") +
  theme_minimal()
```


Predicted probabilities 
```{r}
# Calculate predicted probabilities (response scale)
pred_probs <- emmeans(glmm_full, ~ time_num, type = "response",
                      at = list(time_num = 0:6)) %>%
              as.data.frame()

# Plot
ggplot(pred_probs, aes(x = time_num, y = prob)) +
  geom_line() +
  geom_point() +
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL), alpha = 0.2) +
  scale_y_continuous(limits = c(0,1)) +
  labs(x = "Years since baseline", y = "Predicted probability of CDRSBbin = 1",
       title = "Subject-specific predicted probability over time") +
  theme_minimal()
```




