---
title: "Assignment 3"
author: "Rafke"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 6: Sensitivity Analysis 

```{r}
library(haven)
library(tidyverse)
library(patchwork)
library(pROC)
library(nlme)
library(lme4)
library(data.table)
library(broom)
library(lattice)
library(sjPlot)
library(xtable)
library(kableExtra)
library(mice)
library(miceadds)
library(geepack)
library(mitml)
library(lmerTest)
library(broom.mixed)
library(mitools)
library(zoo) 
library(optimx)
library(dfoptim)
library(mitml)
```


```{r}
dat <- read_sas("./data/alzheimer25.sas7bdat")
str(dat)
```

# Data Preparation
```{r reshape}

reshape_dat <- function(dat) {
    dat_long <- dat %>%
        pivot_longer(
            cols = matches("^(bprs|cdrsb|abpet|taupet)\\d+"),
            names_to = c(".value", "time"),
            names_pattern = "(.+)(\\d+)"
        )
    dat_long
}

to_factor <- function(dat_long) {

    dat_long <- dat_long %>%
        mutate(
            time = as.factor(time),  # coded 0-6
            sex = as.factor(sex),
            edu = as.factor(edu),
            trial = as.factor(trial),
            job = as.factor(job),
            wzc = as.factor(wzc),
            time_num = as.numeric(time)  # coded 1-7
        )
    dat_long
}

dat_long <- reshape_dat(dat)
dat_long <- to_factor(dat_long)

# Baseline per subject (year 1)
base <- dat_long %>%
  filter(time == 0) %>%
  transmute(
    patid, trial,
  )

# Add baseline + time to all rows
dat_long <- dat_long %>%
  left_join(base, by = c("patid","trial"))

rm(base)

dat_long_scaled <- dat_long
vars_to_scale <- c("time_num", "age", "bmi", "adl", "abpet", "taupet", "bprs")
dat_long_scaled[vars_to_scale] <- sapply(dat_long_scaled[vars_to_scale], scale)
```

For the continuous outcome, perform a (brief) sensitivity analysis to assess the potential
impact of MNAR on the substantive conclusions

The LMM model we used in Assignment 1. 
```{r}
lmm_model <- lmer(bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet + (time_num | trial/patid), data = dat_long, REML=FALSE)
summary(lmm_model)
```


Various ways to conduct a sensitivity analysis. There is a consequent pattern in missingness, which can be seen below.

```{r}
colSums(is.na(dat_long))

# 2.1 Create missingness indicator
dat_long_miss <- dat_long %>%
  mutate(missing_bprs = ifelse(is.na(bprs), 1, 0))

# 2.2 Frequency table (Time * Missing)
table(dat_long_miss$time_num, dat_long_miss$missing_bprs)

# 2.3 Proportion missing per time
miss_summary <- dat_long_miss %>%
  group_by(time_num) %>%
  summarise(prop_missing = mean(missing_bprs))

ggplot(miss_summary, aes(x = time_num, y = prop_missing)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::percent(prop_missing, accuracy = 1)),
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, 0.7),
                     breaks = seq(0, 1, 0.1)) + 
  labs(title = "Proportion of missing BPRS by visit year",
       y = "Proportion missing BPRS", x = "Visit year") +
  theme_minimal()
```
### Shift imputed MNAR / delta approach

```{r configuration MI}
# predictors & methods
pred <- quickpred(dat_long, exclude = "patid")
meth <- make.method(dat_long)
meth["bprs"] <- "norm"

# arms to shift
shift_arms <- c("5","12","19")

# delta values to try
delta_grid <- c(-10, -5, 0, 2.5, 5, 10, 15, 20)

m <- 10
maxit <- 20

imp0 <- mice(
  dat_long,
  m = 1,
  method = meth,
  predictorMatrix = pred,
  maxit = 0,
  printFlag = FALSE
)

post_template <- imp0$post
```

```{r warning=FALSE}
results_df <- data.frame(
  delta = numeric(),
  mean_bprs = numeric(),
  sd_bprs = numeric(),
  n_imputed = integer()
)
imp_list <- list()

for (delta in delta_grid) {

  message("Running delta = ", delta)

  post_k <- post_template
  post_k["bprs"] <- paste0(
    "imp[[j]][ where[[i]], i ] <- imp[[j]][ where[[i]], i ] + ",
    "ifelse( as.character(data$trial[ where[[i]] ]) %in% c(",
    paste0("'", shift_arms, "'", collapse = ","),
    "), ", delta, ", 0 )"
  )

  # run MI
  imp_k <- mice(
    dat_long,
    m = m,
    method = meth,
    predictorMatrix = pred,
    post = post_k,
    maxit = maxit,
    printFlag = FALSE
  )
  imp_list[[ as.character(delta) ]] <- imp_k

  d1 <- complete(imp_k, 1)

  results_df <- rbind(
    results_df,
    data.frame(
      delta = delta,
      mean_bprs = mean(d1$bprs, na.rm = TRUE),
      sd_bprs   = sd(d1$bprs, na.rm = TRUE),
      n_imputed = sum(imp_k$where[,"bprs"])
    )
  )
}

```


```{r}
# compare shift-imputed data with the original distribution of BPRS 
example_imp_df <- imp_list[["10"]]
plot(example_imp_df) # check convergence 
densityplot(example_imp_df,    ~ bprs)
stripplot(example_imp_df, bprs ~ .imp)
```

Verify that all NA values were detected and imputed correctly 
```{r}
# imp is a mids object (imp_shift) or you have complete_list
# For mids:
var <- "bprs"
imputed_rows <- which(example_imp_df$where[, var])
length(imputed_rows)                 # how many were imputed
nrow(example_imp_df$imp[[var]])                 # rows in imp$imp[[var]] should match

all(is.na(example_imp_df$data[imputed_rows, var]))

```


```{r}
lmm_model_shift_imp <- lmer(bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet + (time_num | trial/patid), data = example_imp_df$data, REML=FALSE)
summary(lmm_model_shift_imp)
```


A series of MNAR delta-adjusted multiple imputations was performed for the BPRS outcome, varying the delta parameter over a wide and clinically meaningful range (−10 to +20) for the prespecified trial arms. Diagnostic plots indicated good convergence of the imputation algorithm, stable chains across iterations, and plausible imputed distributions. The imputed BPRS values were systematically shifted relative to the observed data in the expected direction, confirming that the delta adjustment was correctly implemented.

Across all delta values, the overall mean and standard deviation of BPRS in the completed datasets remained highly stable. Changes in the delta parameter resulted in only minimal variation in the marginal distribution of BPRS, with no evidence of inflation in variability or implausible distributional behavior. The number of imputed observations was constant across scenarios, and no single delta value produced aberrant or unstable imputations.

Taken together, these findings indicate that this shift-imputation model is well behaved and robust to a broad range of MNAR assumptions. The limited sensitivity of the completed-data distributions to substantial positive and negative delta shifts suggests that the substantive conclusions drawn from subsequent analyses are unlikely to be materially driven by the assumed MNAR mechanism for missing BPRS values.


## Subject-group imputation 

Trial is once again used to compare across multiple (n=25) arms. It is a missingness-mechanism subgroup indicator.
```{r}
# creating the trials arms/subgroups
dat_long$treat <- ifelse(dat_long$trial %in% c("5","12","18"), "1", "0")
dat_long$treat <- factor(dat_long$treat)
# MNAR subgroup (modelobs = treat == '1')
dat_mnar <- dat_long %>% filter(treat == "1")

# Remaining subjects (standard MAR)
dat_mar  <- dat_long %>% filter(treat != "1")
```

```{r}
outcomes <- c("diff4","diff12","diff24","diff52")
m <- 10
blocks_mnar <- as.list(names(dat_mnar))
blocks_mar <- as.list(names(dat_mar))

###### MNAR ########
pred_mnar <- quickpred(dat_mnar, exclude = "patid")
method_mnar <- rep("", length(blocks_mnar))
names(method_mnar) <- names(dat_mnar)
meth_mnar <- make.method(dat_mnar)
for (v in outcomes) {
  blk <- names(blocks_mnar)[
    sapply(blocks_mnar, function(x) v %in% x)
  ]
  meth_mnar[blk] <- "norm"
}

imp_mnar <- mice(
  dat_mnar,
  m = m,
  method = meth_mnar,
  predictorMatrix = pred_mnar,
  blocks = as.list(names(dat_mnar)),
  maxit = 20,
  printFlag = FALSE
)

###### MAR ########
pred_mar <- quickpred(dat_mar, exclude = "patid")
meth_mar <- make.method(dat_mar)
for (v in outcomes) {
  blk <- names(blocks_mar)[
    sapply(blocks_mar, function(x) v %in% x)
  ]
  meth_mar[blk] <- "norm"
}


imp_mar <- mice(
  dat_mar,
  m = m,
  method = meth_mar,
  predictorMatrix = pred_mar,
  blocks = as.list(names(dat_mar)),
  maxit = 20,
  printFlag = FALSE
)

complete_list <- vector("list", m)

for (i in 1:m) {
  c_mnar <- complete(imp_mnar, i)
  c_mar  <- complete(imp_mar,  i)
  
  complete_list[[i]] <- bind_rows(c_mnar, c_mar) %>%
    arrange(patid)
}

pool_all <- function(imp_obj, outcomes) {
  lapply(outcomes, function(v) {
    f <- as.formula(paste(v, "~ treatment + age + sex"))
    pooled <- pool(with(imp_obj, lm(f)))
    summary(pooled)
  }) %>% setNames(outcomes)
}

pooled_mnar <- pool_all(imp_mnar, outcomes)
pooled_mar  <- pool_all(imp_mar, outcomes)
```

This approach preserves subgroup-specific imputation and restores full data structure by recombining MAR & MNAR 

```{r}
mids_combined <- as.mids(complete_list)

# 2) Run model(s) inside the imputation object
# Example model: a single outcome 'diff4' predicted by 'treatment' and covariates
fit_diff4 <- with(mids_combined, lm(diff4 ~ treatment + age + sex))

# 3) Pool the results using Rubin's rules
pooled_diff4 <- pool(fit_diff4)
summary(pooled_diff4)
```


## Different way to do subject-group adjustment

```{r}
formula_analysis <- bprs ~ time_num * (sex + age + adl + job + cdrsb) +
                    wzc + taupet + abpet + (time_num | trial/patid)

ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))

arm_ref <- "1"
dat_long_ref   <- dat_long %>% filter(trial == arm_ref)
dat_long_other <- dat_long %>% filter(trial != arm_ref)

pred_ref   <- quickpred(dat_long_ref, exclude = c("patid"))
pred_other <- quickpred(dat_long_other, exclude = c("patid"))

meth_ref   <- make.method(dat_long_ref);   meth_ref["bprs"] <- "norm"
meth_other <- make.method(dat_long_other); meth_other["bprs"] <- "norm"

m = 10 

imp_ref   <- mice(dat_long_ref,   m = m, method = meth_ref,   predictorMatrix = pred_ref,   maxit = 20, printFlag = FALSE)
imp_other <- mice(dat_long_other, m = m, method = meth_other, predictorMatrix = pred_other, maxit = 20, printFlag = FALSE)

# expand to completed list and stack per-imputation
complete_list <- vector("list", length = m)
for(i in 1:m){
  c_ref   <- complete(imp_ref,   i)
  c_other <- complete(imp_other, i)
  complete_list[[i]] <- bind_rows(c_ref, c_other) %>% arrange(patid, time)
}


# completed_list is a list of length m with completed datasets
m <- length(complete_list)

# fit lmer on each completed dataset
fits_lmer <- lapply(complete_list, function(d) {
  lmer(formula = formula_analysis, data = d, REML = FALSE, control=ctrl)   # REML=FALSE for pooling comparability
})

# extract fixed effects and covariance matrices
coef_list <- lapply(fits_lmer, function(f) as.numeric(fixef(f)))
names_coef <- names(fixef(fits_lmer[[1]]))
V_list    <- lapply(fits_lmer, function(f) {
  as.matrix(vcov(f))                     # variance-covariance of fixed effects
})

# turn into matrices
Qmat <- do.call(cbind, coef_list)        # columns = imputations, rows = coef elements
Qbar <- rowMeans(Qmat)                   # pooled point estimates (vector)

# within-imputation variance Ubar
Ubar <- Reduce("+", V_list) / m          # matrix

# between-imputation variance B
Bmat <- matrix(0, nrow = length(Qbar), ncol = length(Qbar))
for(j in 1:m){
  diffj <- (coef_list[[j]] - Qbar)
  Bmat <- Bmat + tcrossprod(diffj, diffj)
}
Bmat <- Bmat / (m - 1)

# total variance T
Tmat <- Ubar + (1 + 1/m) * Bmat

# pooled SEs and z/t
pooled_se <- sqrt(diag(Tmat))

# Rubin's DF (approx) per Rubin 1987 / Barnard-Rubin
# compute relative increase in variance r = ((1 + 1/m) * diag(B)) / diag(Ubar)
r <- ((1 + 1/m) * diag(Bmat)) / diag(Ubar)
df_old <- (m - 1) * (1 + 1/r)^2            # Rubin's approx; may be Inf if Ubar has zeros

# build result table
res <- data.frame(
  term = names_coef,
  estimate = Qbar,
  se = pooled_se,
  df = df_old,
  t = Qbar / pooled_se,
  p = 2 * pt(-abs(Qbar / pooled_se), df = df_old)
)
rownames(res) <- NULL
res
```
a lot of convergence issues, likely due to the random intercept + slope. Model is likely overparametrized.

```{r}
ctrl <- lmerControl(optCtrl = list(maxfun = 2e5))
fit_try <- lmer(fC, data = complete_list[[8]], REML = FALSE)
af <- allFit(fit_try, maxfun=5e5)  # try out multiple optimizers on the same model formula 
```
find out at which imputation dataset convergence is reached 

```{r}
fail_idx <- which(sapply(fits_lmer, function(x) inherits(x, "try-error") || isSingular(x)))
fail_idx
# inspect that completed dataset
bad_i <- fail_idx[1]
bad_dat <- complete_list[[bad_i]]
```
The maximal random-effects structure led to convergence failures in some imputed datasets; therefore, an uncorrelated random slope–intercept model was experimented with, which converged reliably and yielded comparable fixed-effect estimates.

```{r}
ctrl <- lmerControl(optimizer="bobyqa", optCtrl=list(maxfun=5e5))

# A: uncorrelated slope/intercept per subject
fA <- update(formula_analysis, . ~ . - (time_num | trial/patid) + (time_num || patid))
resA <- try(lmer(fA, data = bad_dat, REML=FALSE, control=ctrl))

# B: random slope+intercept per subject only
fB <- update(formula_analysis, . ~ . - (time_num | trial/patid) + (time_num | patid))
resB <- try(lmer(fB, data = bad_dat, REML=FALSE, control=ctrl))

# C: random intercept only
fC <- update(formula_analysis, . ~ . - (time_num | trial/patid) + (1 | patid))
resC <- try(lmer(fC, data = bad_dat, REML=FALSE, control=ctrl))

list(A = inherits(resA, "merMod") && !isSingular(resA),
     B = inherits(resB, "merMod") && !isSingular(resB),
     C = inherits(resC, "merMod") && !isSingular(resC))
```
```{r}
formula_A <- update(formula_analysis, . ~ . - (time_num | trial/patid) + (time_num || patid))  # A: uncorrelated slope/intercept
formula_B <- update(formula_analysis, . ~ . - (time_num | trial/patid) + (time_num | patid))   # B: slope+intercept per subject
formula_C <- update(formula_analysis, . ~ . - (time_num | trial/patid) + (1 | patid))           # C: intercept only

ctrl <- lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 5e5))

# --- helper: fit list of lmer on complete_list with try-catch ---
fit_on_list <- function(complete_list, formula, control = ctrl){
  fits <- vector("list", length(complete_list))
  for(i in seq_along(complete_list)){
    d <- complete_list[[i]]
    fits[[i]] <- try(lmer(formula, data = d, REML = FALSE, control = control), silent = TRUE)
  }
  return(fits)
}

# --- helper: pool fixed effects using Rubin's rules (returns data.frame) ---
pool_lmer_fixed <- function(fit_list){
  # keep only successful merMod objects
  ok_idx <- which(sapply(fit_list, function(x) inherits(x, "merMod") && !isSingular(x)))
  m_eff  <- length(ok_idx)
  if(m_eff == 0) stop("No successful fits to pool.")
  coef_list <- lapply(fit_list[ok_idx], function(f) as.numeric(fixef(f)))
  names_coef <- names(fixef(fit_list[[ok_idx[1]]]))
  V_list <- lapply(fit_list[ok_idx], function(f) as.matrix(vcov(f)))
  # Qbar
  Qmat <- do.call(cbind, coef_list)      # rows = coef, cols = imputations kept
  Qbar <- rowMeans(Qmat)
  # Ubar
  Ubar <- Reduce("+", V_list) / m_eff
  # B
  Bmat <- matrix(0, nrow = length(Qbar), ncol = length(Qbar))
  for(j in seq_len(m_eff)){
    diffj <- (coef_list[[j]] - Qbar)
    Bmat <- Bmat + tcrossprod(diffj, diffj)
  }
  if(m_eff > 1) Bmat <- Bmat / (m_eff - 1) else Bmat <- Bmat * 0
  # Total variance
  Tmat <- Ubar + (1 + 1/m_eff) * Bmat
  se <- sqrt(diag(Tmat))
  # Rubin DF approx (elementwise)
  r <- ((1 + 1/m_eff) * diag(Bmat)) / diag(Ubar)
  df_old <- (m_eff - 1) * (1 + 1 / r)^2
  df_old[is.nan(df_old) | is.infinite(df_old)] <- NA_real_
  tval <- Qbar / se
  pval <- 2 * pt(-abs(tval), df = df_old)
  res <- data.frame(term = names_coef,
                    estimate = Qbar,
                    se = se,
                    df = df_old,
                    t = tval,
                    p = pval,
                    stringsAsFactors = FALSE)
  attr(res, "m_eff") <- m_eff
  attr(res, "ok_idx") <- ok_idx
  return(res)
}

# --- run fits and pooling for each formula ---
fitsA <- fit_on_list(complete_list, formula_A)
fitsB <- fit_on_list(complete_list, formula_B)
fitsC <- fit_on_list(complete_list, formula_C)

poolA <- try(pool_lmer_fixed(fitsA), silent = FALSE)
poolB <- try(pool_lmer_fixed(fitsB), silent = FALSE)
poolC <- try(pool_lmer_fixed(fitsC), silent = FALSE)

# --- report success counts and show top fixed effects for comparison ---
report_summary <- function(fits, pool_res, label){
  ok <- sum(sapply(fits, function(x) inherits(x,"merMod") && !isSingular(x)))
  total <- length(fits)
  cat(sprintf("%s: %d / %d successful non-singular fits (m_eff = %s)\n", label, ok, total,
              ifelse(inherits(pool_res, "try-error"), "0", attr(pool_res, "m_eff"))))
  if(!inherits(pool_res, "try-error")){
    print(head(pool_res, n = 10))
  } else {
    cat("Pooling failed for", label, "\n")
  }
  cat("\n")
}

report_summary(fitsA, poolA, "Formula A")
report_summary(fitsB, poolB, "Formula B")
report_summary(fitsC, poolC, "Formula C")

# --- compare a specific term of interest across formulas (example: the time*trial interaction terms) ---
# choose terms to compare; change as needed
terms_interest <- c("time_num", grep("time_num:", poolB$term, value = TRUE)[1])  # example
compare_tbl <- data.frame(term = character(0), formula = character(0), estimate = numeric(0), se = numeric(0), stringsAsFactors = FALSE)
for(lst in list(A = poolA, B = poolB, C = poolC)){
  if(!inherits(lst[[1]], "try-error") && !is.null(lst)){
    df <- lst
    df$formula <- deparse(substitute(lst))
    compare_tbl <- bind_rows(compare_tbl, data.frame(term = df$term, formula = deparse(substitute(lst)), estimate = df$estimate, se = df$se, stringsAsFactors = FALSE))
  }
}
# simpler: just print pooled results side-by-side for main terms
cat("Pooled estimates (top terms) for each formula — inspect similarity.\n")
if(!inherits(poolA, "try-error")) print(poolA[1:8, c("term","estimate","se")])
if(!inherits(poolB, "try-error")) print(poolB[1:8, c("term","estimate","se")])
if(!inherits(poolC, "try-error")) print(poolC[1:8, c("term","estimate","se")])

```
