---
title: "Assignment 3"
author: "Rafke"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 6: Sensitivity Analysis 

```{r}
library(haven)
library(tidyverse)
library(patchwork)
library(pROC)
library(nlme)
library(lme4)
library(data.table)
library(broom)
library(lattice)
library(sjPlot)
library(xtable)
library(kableExtra)
library(mice)
library(miceadds)
library(geepack)
library(mitml)
library(lmerTest)
library(broom.mixed)
library(mitools)
library(zoo) 
library(optimx)
library(dfoptim)
library(mitml)
 
```


```{r}
dat <- read_sas("./data/alzheimer25.sas7bdat")
str(dat)
```


# Data Preparation
```{r reshape}

reshape_dat <- function(dat) {
    dat_long <- dat %>%
        pivot_longer(
            cols = matches("^(bprs|cdrsb|abpet|taupet)\\d+"),
            names_to = c(".value", "time"),
            names_pattern = "(.+)(\\d+)"
        )
    dat_long
}

to_factor <- function(dat_long) {

    dat_long <- dat_long %>%
        mutate(
            time = as.factor(time),  # coded 0-6
            sex = as.factor(sex),
            edu = as.factor(edu),
            trial = as.factor(trial),
            job = as.factor(job),
            wzc = as.factor(wzc),
            time_num = as.numeric(time)  # coded 1-7
        )
    dat_long
}

dat_long <- reshape_dat(dat)
dat_long <- to_factor(dat_long)

# Baseline per subject (year 1)
base <- dat_long %>%
  filter(time == 0) %>%
  transmute(
    patid, trial,
  )

# Add baseline + time to all rows
dat_long <- dat_long %>%
  left_join(base, by = c("patid","trial"))

rm(base)

dat_long_scaled <- dat_long
vars_to_scale <- c("time_num", "age", "bmi", "adl", "abpet", "taupet", "bprs")
dat_long_scaled[vars_to_scale] <- sapply(dat_long_scaled[vars_to_scale], scale)
```


For the continuous outcome, perform a (brief) sensitivity analysis to assess the potential
impact of MNAR on the substantive conclusions

The LMM model we used in Assignment 1. 
```{r}
lmm_model <- lmer(bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet + (time_num | trial/patid), data = dat_long_scaled, REML=FALSE)
summary(lmm_model)
```


Various ways to conduct a sensitivity analysis. There is a consequent pattern in missingness in BPRS, which can be seen below.

```{r}
colSums(is.na(dat_long))

# 2.1 Create missingness indicator
dat_long_miss <- dat_long %>%
  mutate(missing_bprs = ifelse(is.na(bprs), 1, 0))

# 2.2 Frequency table (Time * Missing)
table(dat_long_miss$time_num, dat_long_miss$missing_bprs)

# 2.3 Proportion missing per time
miss_summary <- dat_long_miss %>%
  group_by(time_num) %>%
  summarise(prop_missing = mean(missing_bprs))

ggplot(miss_summary, aes(x = time_num, y = prop_missing)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::percent(prop_missing, accuracy = 1)),
            vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, 0.7),
                     breaks = seq(0, 1, 0.1)) + 
  labs(title = "Proportion of missing BPRS by visit year",
       y = "Proportion missing BPRS", x = "Visit year") +
  theme_minimal()
```
### Shift imputed MNAR / delta approach

```{r configuration MI}
# predictors & methods
pred <- quickpred(dat_long, exclude = "patid")
meth <- make.method(dat_long)
meth["bprs"] <- "norm"

# arms to shift
shift_arms <- c("5","12","19")

# delta values to try
delta_grid <- c(5, 10, 15, 20)

m <- 10
maxit <- 20

imp0 <- mice(
  dat_long,
  m = 1,
  method = meth,
  predictorMatrix = pred,
  maxit = 0,
  printFlag = FALSE
)

post_template <- imp0$post
```

```{r warning=FALSE}
results_df <- data.frame(
  delta = numeric(),
  mean_bprs = numeric(),
  sd_bprs = numeric(),
  n_imputed = integer()
)
imp_list <- list()

for (delta in delta_grid) {

  message("Running delta = ", delta)

  post_k <- post_template
  post_k["bprs"] <- paste0(
    "imp[[j]][ where[[i]], i ] <- imp[[j]][ where[[i]], i ] + ",
    "ifelse( as.character(data$trial[ where[[i]] ]) %in% c(",
    paste0("'", shift_arms, "'", collapse = ","),
    "), ", delta, ", 0 )"
  )

  # run MI
  imp_k <- mice(
    dat_long,
    m = m,
    method = meth,
    predictorMatrix = pred,
    post = post_k,
    maxit = maxit,
    printFlag = FALSE
  )
  imp_list[[ as.character(delta) ]] <- imp_k

  d1 <- complete(imp_k, 1)
  # imp_k is mids object for one delta scenario
  fits <- with(
  data = imp_k,
  expr = lmer(
    bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet +
      (time_num | trial/patid),
    REML = FALSE
    )
   )

# Convert mira -> list of fitted models
   fitlist <- fits$analyses

# Pool fixed effects (Rubin’s rules)
   pooled <- testEstimates(fitlist, var.comp = FALSE)
   print(pooled)

  results_df <- rbind(
    results_df,
    data.frame(
      delta = delta,
      mean_bprs = mean(d1$bprs, na.rm = TRUE),
      sd_bprs   = sd(d1$bprs, na.rm = TRUE),
      n_imputed = sum(imp_k$where[,"bprs"])
    )
  )
}

```

## Tipping point to pool LMM across imputations
```{r}
get_mitml_estimates <- function(pooled_obj) {
  # Different mitml versions store the printed table in different slots.
  # We check the common ones.
  candidates <- c("estimates", "results", "table", "coef", "coefficients")

  for (nm in candidates) {
    if (!is.null(pooled_obj[[nm]])) {
      tab <- pooled_obj[[nm]]
      tab <- as.data.frame(tab)
      tab$term <- rownames(tab)
      rownames(tab) <- NULL

      # Ensure required columns exist (your printed output uses these names)
      if (all(c("Estimate", "Std.Error") %in% names(tab))) {
        return(tab)
      }
    }
  }

  # If none found, show structure to help debugging
  stop("Could not extract estimates from pooled_obj. Try str(pooled_obj) to see available slots.")
}

# ---------- Pool one delta ----------
pool_one_delta <- function(mids_obj, delta_value, terms_keep = NULL) {

  fits <- suppressWarnings(with(
    data = mids_obj,
    expr = lmer(
      bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet +
        (time_num | trial/patid),
      REML = FALSE,
      control = lmerControl(
        optimizer = "bobyqa",
        optCtrl = list(maxfun = 2e5),
        check.conv.singular = "ignore"
      )
    )
  ))

  fitlist <- fits$analyses

  pooled <- testEstimates(fitlist, extra.pars = FALSE)

  tab <- get_mitml_estimates(pooled) %>%
    mutate(
      delta = delta_value,
      conf.low  = Estimate - 1.96 * Std.Error,
      conf.high = Estimate + 1.96 * Std.Error,
      n_models = length(fitlist),
      n_singular = sum(sapply(fitlist, isSingular, tol = 1e-4))
    )

  if (!is.null(terms_keep)) tab <- tab %>% filter(term %in% terms_keep)
  tab
}

# ---------- Deltas + term selection ----------
delta_vals <- sort(as.numeric(names(imp_list)))
first_delta <- as.character(delta_vals[1])

tmp_fit <- suppressWarnings(lmer(
  bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet +
    (time_num | trial/patid),
  data = complete(imp_list[[first_delta]], 1),
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 2e5),
    check.conv.singular = "ignore"
  )
))

coef_names <- names(fixef(tmp_fit))

# Keep time effect + any trial-by-time interactions IF they exist in your parametrization
terms_of_interest <- unique(c(
  "time_num",
  grep("time_num:trial|trial.*:time_num", coef_names, value = TRUE)
))

# If you have no trial*time interactions in the model output, don't facet empty panels:
if (length(terms_of_interest) == 1) {
  message("No trial×time interaction terms found in fixef(). Plotting only time_num.")
}

# ---------- Pool across all deltas ----------
pooled_all <- bind_rows(lapply(delta_vals, function(d) {
  pool_one_delta(
    mids_obj = imp_list[[as.character(d)]],
    delta_value = d,
    terms_keep = terms_of_interest
  )
}))

# ---------- Tipping-point figure ----------
ggplot(pooled_all, aes(x = delta, y = Estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ term, scales = "free_y") +
  labs(
    x = "Shift MI applied to imputed BPRS",
    y = "Pooled fixed-effect estimate (95% CI)",
    title = "Pooled fixed-effect estimates under 𝛿-adjusted MI"
  ) +
  theme_minimal()

tipping <- pooled_all %>%
  group_by(term) %>%
  arrange(delta) %>%
  summarise(
    any_cross = any(conf.low <= 0 & conf.high >= 0),
    first_cross_delta = if (any_cross) delta[which(conf.low <= 0 & conf.high >= 0)[1]] else NA_real_,
    .groups = "drop"
  )

print(tipping) # results in 0 since there is no tipping point 

```




## Diagnostic plots for Delta-shift 

```{r}
# compare shift-imputed data with the original distribution of BPRS 
example_imp_df <- imp_list[["10"]]
plot(example_imp_df) # check convergence 
densityplot(example_imp_df,    ~ bprs)
stripplot(example_imp_df, bprs ~ .imp)
```

Verify that all NA values were detected and imputed correctly 
```{r}
# imp is a mids object (imp_shift) or there is complete_list
# For mids:
var <- "bprs"
imputed_rows <- which(example_imp_df$where[, var])
length(imputed_rows)                 # how many were imputed
nrow(example_imp_df$imp[[var]])                 # rows in imp$imp[[var]] should match

all(is.na(example_imp_df$data[imputed_rows, var]))

```


```{r}
lmm_model_shift_imp <- lmer(bprs ~ time_num * (sex + age + adl + job + cdrsb) + wzc + taupet + abpet + (time_num | trial/patid), data = example_imp_df$data, REML=FALSE)
summary(lmm_model_shift_imp)
```
